Understanding Feedback Mechanisms in Machine Learning Jupyter Notebooks

### 2025-03-10 report/emse24-revised.tex
chore: update EMSE response document

\begin{abstract}
  The machine learning development lifecycle is characterized by iterative and exploratory processes that rely on feedback mechanisms to ensure data and model integrity. Despite the critical role of feedback in machine learning engineering, no prior research has been conducted to identify and understand these mechanisms. To address this knowledge gap, we mine 297.8 thousand Jupyter notebooks and analyse 2.3 million code cells. We identify three key feedback mechanisms---assertions, print statements and last cell statements---and further categorize them into implicit and explicit forms of feedback. Our findings reveal extensive use of implicit feedback for critical design decisions and the relatively limited adoption of explicit feedback mechanisms. By conducting detailed case studies with selected feedback instances, we uncover the potential for automated validation of critical assumptions in ML workflows using assertions. Finally, this study underscores the need for improved documentation, and provides practical recommendations on how existing feedback mechanisms in the ML development workflow can be effectively used to mitigate technical debt and enhance reproducibility.
\end{abstract}

@@

\section{Introduction}
The machine learning development lifecycle (MLDL) encompasses a multifaceted series of entangled stages. Significant effort is invested in gathering and assembling the required data from diverse sources. The data then undergoes cleaning and preprocessing to ensure high quality and usability, in the subsequent stages. MLDL is dominated by the model development phase which is highly experimental and iterative. This stage involves exploratory data analysis (EDA) to uncover underlying patterns, fitting the data to various models, and rigorously analysing the results. Often, this stage may lead to additional rounds of feature engineering, where new attributes are created based on insights gained from initial models to enhance model performance. Practitioners may also engage in comparing and contrasting different models to identify the most effective approach, followed by fine-tuning selected models to optimize performance. The final model is then deployed into production where it is continuously monitored, and periodically retrained to adapt to new data or changing conditions~\citep{haakman2021ai,amershi2019software,sculley2015hidden}.

@@

Notebook users can obtain feedback using implicit or explicit mechanisms. Implicit feedback mechanisms require manual validation by the user. For instance, a common prerequisite for all ML models is that the features in the dataset are all numerical. This can be verified with a print statement in a code cell and manually checking the output. This code statement however, will continue to execute even if the data type of a feature changes in subsequent batches of training data.

In contrast, explicit feedback mechanisms document the expected conditions of the code and halt execution if these conditions are not met. For instance, the implicit validation method mentioned above can be replaced with an explicit assertion: \lstinline{assert all(df[col].dtype in ['int64', 'float64'] for col in df.columns), "Not all columns are of numerical type"}. This assert statement will stop the execution if the condition is not satisfied and provide immediate feedback with a clear message explaining the issue.

@@

To the best of our knowledge, no prior studies have been conducted to understand how feedback mechanisms are used by practitioners when developing ML pipelines inside Jupyter notebooks. To address this gap, we conduct a series of empirical studies to understand the role of various feedback mechanisms within Jupyter notebooks. We focus on implicit form of feedback obtained from print statements and output generated by the last statement of code cells, as well as explicit form of feedback obtained from assertions. The research questions along with the contributions of this study are presented below.

@@

    We mine 297.8 thousand Jupyter notebooks written in Python from GitHub and Kaggle. We analyse 2.3 million code cells obtained from the notebooks and identify three key feedback mechanisms used in ML notebooks---explicit feedback from assertions and implicit feedback from print and last cell statements.

@@

    To gain a deeper understanding of Python assertions written in Jupyter notebooks, we conduct individual case-studies with 82 assert statements. We analyse the assertion along with the surrounding code to understand its purpose. Additionally, we analyse the entire code cell, the previous and next cells along with the purpose of the notebook to bring in rich context.

@@

    Similarly, we conduct individual case-studies with 44 print statements and 27 last cell statements. During each case-study, we analyse the code statement along with the raw output to understand its purpose.
\end{description}

Our findings indicate that implicit feedback mechanisms are extensively utilized for making critical design and implementation decisions during the MLDL. While manual validation of these outputs is common, it is also prone to inconsistencies thus highlighting the need for automated validation mechanisms. Assertions are identified as a potential solution, capable of verifying data distributions and ensuring model performance metrics meet predefined thresholds. However, the study also notes the limited use of assertions, with only 8.5\% of the notebooks analysed incorporating them, suggesting a significant gap in current testing practices within the ML community. The findings underscore the necessity for developing ML-specific testing tools that are deeply integrated into the ML development workflow to ensure consistent and reliable validation processes.

@@

A Read-Eval-Print Loop (REPL) is an interactive programming environment that takes single user inputs (Read), executes them (Eval), returns the result to the user (Print) and then waits for the next command (Loop). REPLs offer an immediate feedback loop for quickly testing and debugging code snippets. The Python standard distribution includes a Python REPL with a command line interface to execute Python code in real-time. IPython---a more advanced alternative to the Python REPL---extends the capabilities of the standard Python shell with additional features such as powerful introspection, additional shell syntax, tab completion, and command history. Jupyter notebooks (and its successor Jupyter lab) leverage the IPython kernel to execute Python code and extends its capabilities with a graphical user interface (GUI). Resembling the literate programming paradigm~\citep{knuth1984literate}, Jupyter notebooks offer a web-based interface where users can create and share documents that contain live code, equations, visualizations, and narrative text.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.75\linewidth]{data-collection.pdf}
  \caption{Overview of data collection methodology used in this study}
  \label{fig:data-collection}
\end{figure}

@@

@@

Since the focus of this study is on the analysis of Python code, we only include notebooks that have a valid underlying JSON structure and contain at least one code cell.  Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import at least one of the following popular ML libraries---Scikit Learn~\citep{pedregosa2011scikit-learn}, Pytorch~\citep{paszke2017automatic}, Tensorflow~\citep{abadi2015tensorflow} and Keras~\citep{chollet2015keras}.  The libraries were chosen based on the import analysis conducted by \citet{psallidas2019data}, which analyzed six million Jupyter notebooks collected from GitHub. It is important to note that we excluded libraries such as Pandas, NumPy, and SciPy, since can be utilized for a wide range of tasks beyond machine learning and do not specifically facilitate model training.

We use the Python \lstinline{ast} module to extract the assert, print and last statements from all code cells. Prior to conducting the analysis for RQ2 and RQ3, we remove duplicate data points resulting in a data set of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last cell statements.

\subsection{Exploratory Case Studies}

For RQ2 and RQ3, we allocated a fixed time resource of 200 hours to conduct the exploratory case study analysis of all candidate assertions, print statements and last cell statements. To surface interesting candidates for the case studies, we apply text processing techniques as described below.

@@

% TODO: update this paragraph, candidates were annotated by first and second author
% we calculated IRA and obtained high agreement...
During each case study, we analyse the code of the candidate to understand its purpose. Additionally, we analyse the entire code cell, the previous cell, next cell and the notebook's purpose to bring in rich context. For case studies where the first author was unable to determine the purpose of the candidate, it was additionally analysed by the second and third authors. Further discussions were held with all authors until a consensus was reached.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{notebook-size-distribution.pdf}
  \caption{Number of notebooks in each category of...}
  \label{fig:notebook-size-distribution}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{cell-per-notebook-size-distribution.pdf}
  \caption{Distribution of number of cells for various notebook sizes.}
  \label{fig:cell-per-notebook-size-distribution.pdf}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{fms-per-notebook-size-distribution.pdf}
  \caption{Distribution of assertions, prints and last cell statements for various notebook sizes.}
  \label{fig:fms-per-notebook-size-distribution}
\end{figure}

% TODO should we add the descriptive stats as a table?
% TODO do we need to extrapolate more from figure 5?
Figure~\ref{fig:notebook-size-distribution} shows the distribution of the number of code cells for all notebooks analyzed in this study. The distribution is grouped into three categories---namely small, medium and large---based on the first and third quartiles. This is done since the data is large, non-uniform and contains many outliers. Figure~\ref{fig:cell-per-notebook-size-distribution.pdf} further magnifies the distribution of the number of cells within each category. The dataset mostly contains medium size notebooks (9 to 26 code cells), followed by small (1 to 8 code cells) and large (27 to 772 code cells) size notebooks.

@@

\highlight{\textbf{Observation 1.1} 25 thousand (8.5\%) notebooks contained at least one assertion. 95\% of the 89.6 thousand assertions were written using the built-in Python \lstinline{assert} statement and 5\% using external testing libraries. Methods such as \lstinline{assert_almost_equal} and \lstinline{assert_array_equal} from the numpy and pandas libraries were frequently used to accomodate for the stochastic nature of ML algorithms and comparing high-dimensional data structures.}

@@

\begin{table}
  \centering
  \caption{Examples of the most common tests performed in \lstinline{assert} statements.}
  \begin{tabular}{@{}m{0.5\textwidth} m{0.5\textwidth}@{}}
    \toprule
    \emph{\textbf{Comparison Check}}&
    \emph{\textbf{Function Call}}\\
    \midrule

    \lstinline[]$type(train_dataset[0]) == tuple$&
    \lstinline[]$(np.unique(y_test) == [-1, 1]).all()$\\

    \lstinline[]$torch.__version__ > '1.10.0'$&
    \lstinline[]$isinstance(transmat, pd.dataframe)$\\

    \lstinline[]$len(x_dev) == len(y_dev)$&
    \lstinline[]$np.allclose(np.linalg.norm(wio, axis=0), 1)$\\
    \bottomrule
  \end{tabular}
  \label{tab:common-assert-test}
\end{table}

@@

Of the 95\% assert statements, 28.4 thousand (31.7\%) have a failure message. Figure~\ref{fig:common-assert-msgs} shows the most frequent bi-grams that appear in these failure messages. Similar to the analysis presented above, we removed the duplicate messages prior to analysing the bi-grams resulting in 5.9 thousand unique messages. The bi-grams indicate that the shape and data type of data structures are frequently verified.

@@

We collected 1.4 million print statements from 180 thousand notebooks (60\% of the total notebooks analysed in this study). Unlike assert statements, print statements do not have a fixed lexical semantic since the Python \lstinline{ast} module interprets them as regular function calls. Additionally, print statements can be written in several forms---the \lstinline{print} function may have multiple arguments, any form of string interpolation supported by Python or output from function calls. Figure~\ref{fig:common-print-nodes} shows that the most common AST nodes inside of print statements are constants followed by function calls. Figure~\ref{fig:common-print-constant-types} further validates that the constants are almost entirely of type string. Based on these observations, we dissect the print statements into two components---text written in natural language (from the \emph{Constant} nodes) and code (from all other nodes). Figure~\ref{fig:common-print-constants} and Figure~\ref{fig:common-print-not-constants} present the most common bi-grams that appear in the natural text and code of print statements respectively. Both figures collectively indicate that print statements are frequently used to print various performance metrics of ML models.

@@

We collected 1 million last cell statements from 119 thousand notebooks (40\% of the total notebooks analysed in this study). Figure~\ref{fig:common-last-nodes} shows that \emph{Call} nodes frequently occur in the AST of last cell statements. Figure~\ref{fig:common-last-modules} shows the most common Python modules and objects while Figure~\ref{fig:common-last-functions} show the most common functions used inside these \emph{Call} nodes. The figures highlight the prominence of data visualization libraries seaborn (\lstinline{sns}) and matplotlib (\lstinline{plt}) and visualization functions such as \lstinline{plot} and \lstinline{countplot} in the notebooks. The data analysis tool pandas (represented by \lstinline{df} and \lstinline{pd}) also features prominently with many data exploration and preprocessing functions such as \lstinline{head}, \lstinline{value_counts} and \lstinline{describe} being used frequently. The presence of \lstinline{train} and \lstinline{model} Python objects along with the \lstinline{fit} function indicates the use of ML libraries.

@@

We further analysed the most common functions called on pandas dataframes by isolating the last cell statements that contain \lstinline{data} or \lstinline{df} objects. This is shown in Figure~\ref{fig:common-last-df-functions} which reveals a focus on exploratory data analysis and statistical operations. The frequent use of functions such as \lstinline{value_counts}, \lstinline{head}, and \lstinline{unique}, indicate an emphasis on understanding the structure and content of the data. Visualization functions like \lstinline{plot} and \lstinline{hist} suggest that practitioners often create quick visualizations directly from dataframes. Descriptive statistics are also prominent, with functions like \lstinline{describe}, \lstinline{mean}, and \lstinline{sum} being commonly used.

Similarly, we also analysed the most common visualization functions used in notebooks by isolating the last cell statements where the module is \lstinline{sns} or \lstinline{plt}. This is shown in Figure~\ref{fig:common-last-visualization-functions} which indicates that categorical data visualization appears to be a primary focus, with \lstinline{countplot} and \lstinline{barplot} ranking high among the most frequently used functions. The prominence of \lstinline{heatmap} and \lstinline{boxplot} functions indicates a regular need for visualizing correlations and statistical distributions. For continuous data, functions like \lstinline{distplot}, \lstinline{scatterplot}, and \lstinline{scatter} are commonly used. The presence of \lstinline{imshow} suggests that image-related visualizations are also significant in many notebooks.

@@

\highlight{\textbf{Observation 2.1} 26 assertions verify the dimensions of various data structures such as input features, labels, predictions, images, sequences, and text embeddings. These checks ensure data consistency and compatibility throughout the ML pipeline, preventing runtime errors and ensuring accurate model training and evaluation.}

@@

\highlight{\textbf{Observation 2.2} Fourteen assertions perform validation checks on data structures, ensuring the presence of specific values, verifying value ranges, and confirming uniqueness or cardinality of values. These checks catch potential errors or inconsistencies early, ensuring data meets specific criteria or constraints for robust ML applications.}

@@

This study finds several assertions used to test model performance against predefined thresholds for key metrics such as accuracy, precision, recall and F1 score. Range checks confirm that the performance metrics fall within acceptable boundaries, indicating good predictive performance without overfitting (\emph{A38}). Direct comparisons of model outputs with expected results, validate the model's predictive accuracy and reliability under operational conditions (\emph{A15}). Other assertions ensure that the learned parameters align closely with theoretically or empirically derived values, further cementing the model’s statistical validity (\emph{A19}). Finally, assertions are also used to check for the expected number of neighbours as an indirect measure of model performance in specific scenarios like clustering (\emph{A7}).

@@

\highlight{\textbf{Observation 2.4} Eight assertions ensure the presence of specific columns in the dataset and the absence of missing values. These checks prevent operations on empty datasets and confirm the absence of \texttt{NaN} values, maintaining data integrity and reliability for model training.}

@@

Type check assertions are used to maintain data integrity, ensure compatibility between different components of the ML pipeline, and preventing errors and unexpected behaviour. Type checks are commonly performed to ensure that the input features are PyTorch float tensors or NumPy arrays, a requirement for many deep learning models and scientific computing operations respectively (\emph{A2, A88}). Type checks are also applied to verify the type of the ML models (\emph{A40}), data structures and intermediate objects (\emph{A35}) and individual columns or elements within a dataset (\emph{A81}).

\highlight{\textbf{Observation 2.6} Five assertions validate the data types of input features, models, and intermediate objects. These checks ensure compatibility with ML models and operations, preventing errors from incompatible methods or attributes, thus maintaining the reliability of the ML workflow.}

@@

When working with neural networks and other statistical models, it is imperative to ensure that mathematical properties are preserved throughout operations involving arrays and matrices. This rigour is captured though assertions that ensure convolution operations are dimensionally consistent (\emph{A3}), checking for the symmetry of matrices (\emph{A64}) and monitoring the standard deviation of outputs to decide the appropriateness of using batch normalization (\emph{A25}). These assertions are not just precautionary but are vital for confirming that the underpinnings of ML algorithms align with expected mathematical principles.

\highlight{\textbf{Observation 2.7} Four assertions validate key mathematical properties of data and model outputs. These include ensuring dimensional consistency in convolution operations, monitoring standard deviation for batch normalization suitability, verifying uniformity in computations, and checking matrix symmetry. These checks confirm the robustness and reliability of model computations.}

@@

\highlight{\textbf{Observation 2.8} This study identifies three assertions validating batch size configurations. These assertions ensure batch sizes are divisible by dataset sizes, image dimensions are appropriately divided by patch sizes, and the number of images exceeds the batch size. These checks are crucial for optimizing computational efficiency and ensuring stable and smooth model training.}

@@

\highlight{\textbf{Observation 2.9} Three assertions verify the consistency of neural network architecture. These include ensuring weight dimensions match across layers and that activation shapes and regularization parameters are as expected. These checks ensure that the neural network is correctly configured and integrated with pre-trained models, maintaining the integrity of data flow during training.}

@@

\highlight{\textbf{Observation 2.10} This study identifies one assertion to ensure there are no overlapping values between training and validation sets to prevent data leakage. Preventing data leakage is essential to avoid overfitting and ensure the model's ability to generalize to new data. This practice maintains the integrity and validity of the model evaluation metrics.}

@@

\highlight{\textbf{Observation 3.4} This study finds five implicit spot checks performed at various stages of the MLDL. Spot checks are used to ensure data integrity by verifying data transformation and feature engineering steps meet expectations.}

>>>>

\begin{abstract}
	The machine learning development lifecycle is characterized by iterative and exploratory processes that rely on feedback mechanisms to ensure data and model integrity. Despite the critical role of feedback in machine learning engineering, no prior research has been conducted to identify and understand these mechanisms. To address this knowledge gap, we mine 297.8 thousand Jupyter notebooks and analyze 2.3 million code cells. We identify three key feedback mechanisms---assertions, print statements and last cell statements---and further categorize them into implicit and explicit forms of feedback. Our findings reveal extensive use of implicit feedback for critical design decisions and the relatively limited adoption of explicit feedback mechanisms. By conducting detailed case studies with selected feedback instances, we uncover the potential for automated validation of critical assumptions in ML workflows using assertions. Finally, this study underscores the need for improved documentation, and provides practical recommendations on how existing feedback mechanisms in the ML development workflow can be effectively used to mitigate technical debt and enhance reproducibility.
\end{abstract}

@@

\section{Introduction}
The machine learning development lifecycle (MLDL) encompasses a multifaceted series of entangled stages. Significant effort is invested in gathering and assembling the required data from diverse sources. The data then undergoes cleaning and preprocessing to ensure high quality and usability, in the subsequent stages. MLDL is dominated by the model development phase which is highly experimental and iterative. This stage involves exploratory data analysis (EDA) to uncover underlying patterns, fitting the data to various models, and rigorously analyzing the results. Often, this stage may lead to additional rounds of feature engineering, where new attributes are created based on insights gained from initial models to enhance model performance. Practitioners may also engage in comparing and contrasting different models to identify the most effective approach, followed by fine-tuning selected models to optimize performance. The final model is then deployed into production where it is continuously monitored, and periodically retrained to adapt to new data or changing conditions~\citep{haakman2021ai,amershi2019software,sculley2015hidden}.

@@

Within this environment, feedback mechanisms are essential for improving code quality and diagnosing issues. \textbf{Feedback mechanisms are tools or techniques intentionally integrated into the code by the developer to gain insight into the execution of the program}. Importantly, not all forms of feedback quality as feedback mechanisms. For example, a bug in the code may produce feedback indicating an issue, but because the bug was not intentionally inserted, it does not count as a feedback mechanism.

Notebook users can obtain feedback using implicit or explicit mechanisms. \textbf{Implicit feedback mechanisms require manual validation by the user}. For instance, a common prerequisite for all ML models is that the features in the dataset are all numerical. This can be verified with a print statement in a code cell and manually checking the output. This code statement however, will continue to execute even if the data type of a feature changes in subsequent batches of training data.

In contrast, \textbf{explicit feedback mechanisms document the expected conditions of the code and halt execution if these conditions are not met}. For instance, the implicit validation method mentioned above can be replaced with an explicit assertion: \lstinline{assert all(df[col].dtype in ['int64', 'float64'] for col in df.columns), "Not all columns are of numerical type"}. This assert statement will stop the execution if the condition is not satisfied and provide immediate feedback with a clear message explaining the issue.

It is important to contextualize feedback mechanisms within the broader framework of software testing. Traditional testing methodologies employ formal tests to explicitly identify discrepancies between expected and actual code functionality. \textbf{While all tests serve as feedback mechanisms, not all feedback mechanisms operate within this formal testing paradigm}. For instance, print statements and outputs from the last statement of code cells offer a more informal approach to gaining insights into the execution status and data flow within a Jupyter notebook. While such implicit feedback types can help practitioners diagnose issues within their code, they necessitate manual validation and interpretation, which can increase the potential for oversight and errors.

@@

To the best of our knowledge, no prior studies have been conducted to understand how feedback mechanisms are used by practitioners when developing ML pipelines inside Jupyter notebooks. To address this gap, we conduct a series of empirical studies to understand the role of various feedback mechanisms within Jupyter notebooks. We focus on implicit form of feedback obtained from print statements and output generated by the last statement of code cells, as well as explicit form of feedback obtained from assertions.

The research questions along with the contributions of this study are presented below.

@@

	      We mine 297.8 thousand Jupyter notebooks written in Python from GitHub and Kaggle. We analyze 2.3 million code cells obtained from the notebooks and identify three key feedback mechanisms used in ML notebooks---explicit feedback from assertions and implicit feedback from print and last cell statements.

@@

	      To gain a deeper understanding of Python assertions written in Jupyter notebooks, we conduct individual case-studies with 82 assert statements. We analyze the assertion along with the surrounding code to understand its purpose. Additionally, we analyze the entire code cell, the previous and next cells along with the purpose of the notebook to bring in rich context.

@@

	      Similarly, we conduct individual case-studies with 44 print statements and 27 last cell statements. During each case-study, we analyze the code statement along with the raw output to understand its purpose.
\end{description}

Our findings indicate that implicit feedback mechanisms are extensively utilized for making critical design and implementation decisions during the MLDL. While manual validation of these outputs is common, it is also prone to inconsistencies thus highlighting the need for automated validation mechanisms. Assertions are identified as a potential solution, capable of verifying data distributions and ensuring model performance metrics meet predefined thresholds. However, the study also notes the limited use of assertions, with only 8.5\% of the notebooks analyzed incorporating them, suggesting a significant gap in current testing practices within the ML community. The findings underscore the necessity for developing ML-specific testing tools that are deeply integrated into the ML development workflow to ensure consistent and reliable validation processes.

@@

A Read-Eval-Print Loop (REPL) is an interactive programming environment that takes single user inputs (Read), executes them (Eval), returns the result to the user (Print) and then waits for the next command (Loop). REPLs offer an immediate feedback loop for quickly testing and debugging code snippets. The Python standard distribution includes a Python REPL with a command line interface to execute Python code in real-time. IPython---a more advanced alternative to the Python REPL---extends the capabilities of the standard Python shell with additional features such as powerful introspection, additional shell syntax, tab completion, and command history. Jupyter notebooks (and its successor Jupyter lab) leverage the IPython kernel to execute Python code and extend its capabilities with a graphical user interface (GUI). Resembling the literate programming paradigm~\citep{knuth1984literate}, Jupyter notebooks offer a web-based interface where users can create and share documents that contain live code, equations, visualizations, and narrative text.

@@

% NOTE: above number extracted using the following shell command: `cat data-collection.log |grep '^OUTPUT'|cut -d: -f2|sed -e 's/-asserts\.csv//' -e 's/-prints\.csv//' -e 's/-lasts\.csv//'|sort|uniq|wc -l`
\begin{figure}
	\centering
	\includegraphics[width=0.75\linewidth]{data-collection.pdf}
	\caption{Overview of data collection methodology used in this study}
	\label{fig:data-collection}
\end{figure}

@@

\subsubsection{Collection of ML Jupyter Notebooks}

@@

This study focuses on analyzing Python code in Jupyter notebooks with valid JSON structures containing at least one code cell. To concentrate specifically on machine learning projects, we only included notebooks that import at least one major ML library: Scikit Learn~\citep{pedregosa2011scikit-learn}, PyTorch~\citep{paszke2017automatic}, TensorFlow~\citep{abadi2015tensorflow}, or Keras~\citep{chollet2015keras}. While this approach may exclude some ML projects using alternative libraries, research by \citet{psallidas2019data} examining six million GitHub Jupyter notebooks confirms that most ML-oriented notebooks utilize these frameworks. Despite their widespread usage, we deliberately excluded Pandas, NumPy, and SciPy from our selection criteria as these libraries serve numerous purposes beyond machine learning and are not specifically designed for model training. After applying these filters, we excluded 107 thousand notebooks, resulting in a final dataset of 190.8 thousand notebooks for analysis.

\subsubsection{Collection of Feedback Mechanisms}

For RQ1, we extracted the contents of all code cells from the Jupyter notebooks by parsing the underlying JSON structure. We then used the \lstinline{ast} module provided by the Python programming language to extract the assert, print and last statements from all code cells.

The raw contents of each cell were then converted into an AST. The \lstinline{NodeVisitor} class from the \lstinline{ast} module was used to traverse the AST and collect all \lstinline{Assert} nodes. To collect assertions written using Python modules not included in the standard distribution, we also collected all function \lstinline{Call} nodes that contain the string ``assert'' in their name. Similarly, all print statements were collected by traversing the AST to find all function \lstinline{Call} nodes with the name ``print''. To collect the last statements, we first removed all cells that did not produce any output and consequently extracted the last top-level node from the AST of the code cell. The raw Python code was regenerated from the ASTs using the \lstinline{unparse} method provided by the \lstinline{ast} module.

Prior to conducting the analysis for RQ2 and RQ3, we removed duplicate data points resulting in a data set of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last cell statements.

\subsection{Exploratory Case Study Analysis}

For RQ2 and RQ3, we allocated a fixed time resource of 200 hours to conduct the exploratory case study analysis of all candidate assertions, print statements and last cell statements. We specified Jupyter notebooks written in the Python programming language as the \emph{site} for the study. Given the exploratory nature of this study, we emphasized capturing a diverse range of candidates by apply text processing and sampling techniques as described below.

@@

% NOTE: Cohen's kappa: 0.8083955317337074

During each case study, the first author categorized each candidate using \emph{inductive coding}. The first author analyzed the Python code of the candidate to understand its purpose. Additionally, the entire code cell, the previous cell, next cell and the notebook's purpose was analyzed to bring in rich context. For case studies where the first author was unable to determine the purpose of the candidate, it was additionally analyzed by the second and third authors. Finally, the entire research team held regular meetings throughout this iterative process to ensure consistency in the categorization of the feedback mechanisms.

A taxonomy of feedback mechanisms observed in ML Jupyter notebooks was created using the categories generated during the exploratory case study analysis. To validate the taxonomy developed by the first author, the second author independently annotated the candidates using the categories developed by the first author. The inter-rater agreement was calculated using the \emph{Cohen's Kappa} method and a score 0.8 was obtained which is regarded as highly agreeable.

@@

\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth]{notebook-size-distribution.pdf}
	\caption{Number of notebooks in the small, medium and large notebook size categories.}
	\label{fig:notebook-size-distribution}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.75\textwidth]{cell-per-notebook-size-distribution.pdf}
	\caption{Distribution of number of cells in all notebook size categories.}
	\label{fig:cell-per-notebook-size-distribution}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=0.5\textwidth]{fms-per-notebook-size-distribution.pdf}
	\caption{Distribution of assertions, prints and last cell statements across all notebook size categories.}
	\label{fig:fms-per-notebook-size-distribution}
\end{figure}

Figure~\ref{fig:notebook-size-distribution} shows the distribution of the number of code cells for all notebooks analyzed in this study. The distribution is grouped into three categories---namely small, medium and large---based on the first and third quartiles. This is done since the data is large, non-uniform and contains many outliers. Figure~\ref{fig:cell-per-notebook-size-distribution} further magnifies the distribution of the number of cells within each category. The dataset mostly contains medium size notebooks (9 to 26 code cells), followed by small (1 to 8 code cells) and large (27 to 772 code cells) size notebooks.

@@

\highlight{\textbf{Observation 1.1} 25 thousand (8.5\%) notebooks contained at least one assertion. 95\% of the 89.6 thousand assertions were written using the built-in Python \lstinline{assert} statement and 5\% using external testing libraries. Methods such as \lstinline{assert_almost_equal} and \lstinline{assert_array_equal} from the numpy and pandas libraries were frequently used to accommodate for the stochastic nature of ML algorithms and to compare high-dimensional data structures.}

@@

\begin{table}
	\centering
	\caption{Examples of the most common tests performed in \lstinline{assert} statements.}
	\begin{tabular}{@{}m{0.5\textwidth} m{0.5\textwidth}@{}}
		\toprule
		\emph{\textbf{Comparison Check}}              &
		\emph{\textbf{Function Call}}                             \\
		\midrule

		\lstinline[]$type(train_dataset[0]) == tuple$ &
		\lstinline[]$(np.unique(y_test) == [-1, 1]).all()$        \\

		\lstinline[]$torch.__version__ > '1.10.0'$    &
		\lstinline[]$isinstance(transmat, pd.dataframe)$          \\

		\lstinline[]$len(x_dev) == len(y_dev)$        &
		\lstinline[]$np.allclose(np.linalg.norm(wio, axis=0), 1)$ \\
		\bottomrule
	\end{tabular}
	\label{tab:common-assert-test}
\end{table}

@@

Of the 95\% assert statements, 28.4 thousand (31.7\%) have a failure message. Figure~\ref{fig:common-assert-msgs} shows the most frequent bi-grams that appear in these failure messages. Similar to the analysis presented above, we removed the duplicate messages prior to analyzing the bi-grams resulting in 5.9 thousand unique messages. The bi-grams indicate that the shape and data type of data structures are frequently verified.

@@

We collected 1.4 million print statements from 180 thousand notebooks (60\% of the total notebooks analyzed in this study). Unlike assert statements, print statements do not have a fixed lexical semantic since the Python \lstinline{ast} module interprets them as regular function calls. Additionally, print statements can be written in several forms---the \lstinline{print} function may have multiple arguments, any form of string interpolation supported by Python or output from function calls. Figure~\ref{fig:common-print-nodes} shows that the most common AST nodes inside print statements are constants followed by function calls. Figure~\ref{fig:common-print-constant-types} further validates that the constants are almost entirely of type string. Based on these observations, we dissect the print statements into two components---text written in natural language (from the \emph{Constant} nodes) and code (from all other nodes). Figure~\ref{fig:common-print-constants} and Figure~\ref{fig:common-print-not-constants} present the most common bi-grams that appear in the natural text and code of print statements respectively. Both figures collectively indicate that print statements are frequently used to print various performance metrics of ML models.

@@

We collected 1 million last cell statements from 119 thousand notebooks (40\% of the total notebooks analyzed in this study). Figure~\ref{fig:common-last-nodes} shows that \emph{Call} nodes frequently occur in the AST of last cell statements. Figure~\ref{fig:common-last-modules} shows the most common Python modules and objects while Figure~\ref{fig:common-last-functions} shows the most common functions used inside these \emph{Call} nodes. The figures highlight the prominence of data visualization libraries seaborn (\lstinline{sns}) and matplotlib (\lstinline{plt}) and visualization functions such as \lstinline{plot} and \lstinline{countplot} in the notebooks. The data analysis tool pandas (represented by \lstinline{df} and \lstinline{pd}) also features prominently with many data exploration and preprocessing functions such as \lstinline{head}, \lstinline{value_counts} and \lstinline{describe} being used frequently. The presence of \lstinline{train} and \lstinline{model} Python objects along with the \lstinline{fit} function indicates the use of ML libraries.

@@

We further analyzed the most common functions called on pandas dataframes by isolating the last cell statements that contain \lstinline{data} or \lstinline{df} objects. This is shown in Figure~\ref{fig:common-last-df-functions} which reveals a focus on exploratory data analysis and statistical operations. The frequent use of functions such as \lstinline{value_counts}, \lstinline{head}, and \lstinline{unique}, indicate an emphasis on understanding the structure and content of the data. Visualization functions like \lstinline{plot} and \lstinline{hist} suggest that practitioners often create quick visualizations directly from dataframes. Descriptive statistics are also prominent, with functions like \lstinline{describe}, \lstinline{mean}, and \lstinline{sum} being commonly used.

Similarly, we also analyzed the most common visualization functions used in notebooks by isolating the last cell statements where the module is \lstinline{sns} or \lstinline{plt}. This is shown in Figure~\ref{fig:common-last-visualization-functions} which indicates that categorical data visualization appears to be a primary focus, with \lstinline{countplot} and \lstinline{barplot} ranking high among the most frequently used functions. The prominence of \lstinline{heatmap} and \lstinline{boxplot} functions indicates a regular need for visualizing correlations and statistical distributions. For continuous data, functions like \lstinline{distplot}, \lstinline{scatterplot}, and \lstinline{scatter} are commonly used. The presence of \lstinline{imshow} suggests that image-related visualizations are also significant in many notebooks.

@@

\highlight{\textbf{Observation 2.1} 26 assertions verify the dimensions of various data structures to ensure data consistency and compatibility throughout the ML pipeline, prevent runtime errors and ensure accurate model training and evaluation.}

@@

\highlight{\textbf{Observation 2.2} Fourteen assertions perform validation checks on data structures to catch errors or inconsistencies early, ensuring data meets specific criteria or constraints for robust ML applications.}

@@

This study finds several assertions used to test model performance against predefined thresholds for key metrics such as accuracy, precision, recall and F1 score. Range checks confirm that the performance metrics fall within acceptable boundaries, indicating good predictive performance without overfitting (\emph{A38}). Direct comparisons of model outputs with expected results, validate the model's predictive accuracy and reliability under operational conditions (\emph{A15}). Other assertions ensure that the learned parameters align closely with theoretically or empirically derived values, further cementing the model’s statistical validity (\emph{A19}). Finally, assertions are also used to check for the expected number of neighbors as an indirect measure of model performance in specific scenarios like clustering (\emph{A7}).

@@

\highlight{\textbf{Observation 2.4} Eight assertions ensure the presence of specific columns in the dataset and the absence of \texttt{NaN} or missing values. These checks prevent operations on empty datasets to maintain data integrity and reliability for model training.}

@@

Type check assertions are used to maintain data integrity, ensure compatibility between different components of the ML pipeline, and preventing errors and unexpected behavior. Type checks are commonly performed to ensure that the input features are PyTorch float tensors or NumPy arrays, a requirement for many deep learning models and scientific computing operations respectively (\emph{A2, A88}). Type checks are also applied to verify the type of the ML models (\emph{A40}), data structures and intermediate objects (\emph{A35}) and individual columns or elements within a dataset (\emph{A81}).

\highlight{\textbf{Observation 2.6} Five assertions validate the data types of input features and Python objects. These checks prevent errors from incompatible methods or attributes, and improve the reliability of the ML workflow.}

@@

When working with neural networks and other statistical models, it is imperative to ensure that mathematical properties are preserved throughout operations involving arrays and matrices. This rigor is captured though assertions that ensure convolution operations are dimensionally consistent (\emph{A3}), checking for the symmetry of matrices (\emph{A64}) and monitoring the standard deviation of outputs to decide the appropriateness of using batch normalization (\emph{A25}). These assertions are not just precautionary but are vital for confirming that the underpinnings of ML algorithms align with expected mathematical principles.

\highlight{\textbf{Observation 2.7} Four assertions validate key mathematical properties of data and model outputs to confirm the robustness and reliability of model computations.}

@@

\highlight{\textbf{Observation 2.8} Three assertions validate the batch size configurations. Such checks are crucial for optimizing computational efficiency and ensuring stable and smooth model training.}

@@

\highlight{\textbf{Observation 2.9} Three assertions verify that the neural network is correctly configured and integrated with pre-trained models to maintain the integrity of data flow during training.}

@@

\highlight{\textbf{Observation 2.10} One assertion ensures there are no overlapping values between the training and validation sets to prevent data leakage and subsequently overfitting during model training.}

@@

\highlight{\textbf{Observation 3.4} Five implicit spot checks are performed throughout the various stages of the MLDL to verify data transformation and feature engineering steps meet expectations.}
